### 2018-05-07 1_intro.tex
add a bit up to hooks.

Traditionally, studies related to concurrent programming concern on more fundamental theoretical questions of designing race-free and lock-free parallel algorithms, asynchronous data structures and synchronisation primitives of a programming language.
Unfortunately, when it comes to the real-world concurrent programs, the algorithmic level of abstraction is not enough for guaranteeing their properties of correctness and reliability.
The reasons of this fact lie in the code optimisations that both compiler and hardware perform in order to increase performance as much as possible.
For instance, Figure~\ref{simple_wmm_x86} provides simple example of reachability of the state `\texttt{(0:EAX=0~/\textbackslash~1:EAX=0)}' on the x86 architecture (such little examples that illustrate specific behaviour of a WMM are called \textit{litmus tests}).
%reordering of memory access instructions within single process allowed by the x86-TSO weak memory model, which potentially breaks the program logic. 
This state is allowed because in the x86 architecture each processor may cache the write to shared memory variable into its local write buffer, so that they do not immediately become visible by processes running on other cores.
In the example, the write `\texttt{MOV~[x],1}' performed by process \texttt{P0} stores value~\texttt{1} to the shared variable~\texttt{[x]} into the write buffer of process~\texttt{P0}.
Meanwhile, the write cache of the process \texttt{P1} may not have an updated version of the variable~\texttt{[x]}, neither the main memory, so that the read `\texttt{MOV~EBX,[x]}' performed in the process~\texttt{P1} may read the initial value~\texttt{0} even if this variable has been already updated in the other thread.
These problems have lead to the need for formalisation of semantics of memory operations within different concurrent architectures defined by \textit{weak memory models (WMMs)}.

@@

Research of weak memory models firstly aims to \textit{formalise} develop the formal approach of understanding programs with respect to weak memory models which is systematic, sound and complete. The first (and so far the only) such a framework was presented in 2010~\cite{alglave2010shared}.
In addition to developing rather theoretical basis, researchers work on extracting the WMMs for hardware architectures from existing implementations of from their specifications, which are written in natural language and thus suffer from ambiguities and incompleteness.
Over last decade the memory models have been defined for most mainstream multiprocessor architectures, such as x86-TSO and Sparc-TSO (for \textit{Total Store Order}) model for x86 and Sparc architecture formalised in 2009~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding, alglave2014herding}, and others.
There are projects for validating hardware architectures wrt. a memory model, e.g.~\cite{lustig2014pipecheck,lustig2016coatcheck}.

>>>>

Traditionally, studies related to concurrent programming concern on more fundamental theoretical questions of designing race-free and lock-free parallel algorithms, asynchronous data structures and synchronisation primitives of a programming language.
Unfortunately, when it comes to the real-world concurrent programs, the algorithmic level of abstraction is not enough for guaranteeing their properties of correctness and reliability.
The reasons of this fact lie in the code optimisations that both compiler and hardware perform in order to increase performance as much as possible~\cite{adve1996shared}.
For instance, Figure~\ref{simple_wmm_x86} provides simple example of reachability of the state `\texttt{(0:EAX=0~/\textbackslash~1:EAX=0)}' on the x86 architecture (such little examples that illustrate specific behaviour of a WMM are called \textit{litmus tests}).
%reordering of memory access instructions within single process allowed by the x86-TSO weak memory model, which potentially breaks the program logic. 
This state is allowed because in the x86 architecture each processor may cache the write to shared memory variable into its local write buffer, so that they do not immediately become visible by processes running on other cores.
In the example, the write `\texttt{MOV~[x],1}' performed by process \texttt{P0} stores value~\texttt{1} to the shared variable~\texttt{[x]} into the write buffer of process~\texttt{P0}.
Meanwhile, the write cache of the process \texttt{P1} may not have an updated version of the variable~\texttt{[x]}, neither the main memory, so that the read `\texttt{MOV~EBX,[x]}' performed in the process~\texttt{P1} may read the initial value~\texttt{0} even if this variable has been already updated in the other thread.
These problems have lead to the need for formalisation of semantics of memory operations within different concurrent architectures defined by \textit{weak memory models (WMMs)}.

@@

Research of weak memory models firstly aims to \textit{formalise} develop the formal approach of understanding programs with respect to weak memory models which is systematic, sound and complete. The first (and so far the only) such a framework was formalised by Jade Alglave in 2010~\cite{alglave2010shared}.
In addition to developing rather theoretical basis, researchers work on extracting the WMMs for hardware architectures from existing implementations of from their specifications, which are written in natural language and thus suffer from ambiguities and incompleteness.
Over last decade the memory models have been defined for most mainstream multiprocessor architectures, such as x86-TSO and Sparc-TSO (for \textit{Total Store Order}) model for x86 and Sparc architecture formalised in 2009~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding, alglave2014herding}, and others.
There are projects for validating hardware architectures wrt. a memory model, e.g.~\cite{lustig2014pipecheck,lustig2016coatcheck}.

### 2018-05-07 3_encod.tex
add a bit up to hooks.

%TODO: more about SMT (= SAT + constraints expressed in background theories. Exapmles of theories that we use and that are implem-ted in Z3).

As it has been discussed in Chapter~\ref{ch:intro}, the program may behave differently when compiled for different parallel hardware architectures. This may cause the portability bugs, the behaviour that is allowed under one architecture and forbidden under another. 
%already said:
%Although the research of weak memory models has achieved considerable success, most works remain to be rather theoretical that practical and serve mostly as tools for better understanding the concurrent nature of programs.
% The first tool that tackles the problem 
%The first work that has investigated the practical approach of modeling and verification the real-world programs with respect to weak memory models was ~\cite{Porthos17SAS}  
In this Chapter, we describe the general task of analysing the concurrent software portability
%may be stated 
as a \textit{bounded reachability} problem, which in turn can be reduced to a SAT problem~\cite{Porthos17SAS} (more precisely, to an SMT problem).

@@

Note that the definition of portability requirements against \textit{executions} is strong enough, as it implies the portability against \textit{states} (the \textit{state-portability})~\cite{Porthos17CoRR}.
The result SMT formula $\phi$ that encodes the portability problem should contain both encodings of control-flow $\phi_{CF}$ and data-flow $\phi_{DF}$ of the program, and assertions of both memory models: ${\phi = \phi_{CF} \land \phi_{DF} \land \phi_{\mathcal{M_T}} \land \phi_{\lnot\mathcal{M_S}}}$. If the formula is satisfiable, there exist a portability bug.
%The control-flow and data-flow encodings are standard for BMC~\cite{collavizza2006exploration}, they are described below. 
%However, encoding of memory models requires additional techniques due to recursive definitions of relations, that were proposed in~\cite{Porthos17SAS}.

>>>>

%TODO: more about SMT-(= SAT + constraints expressed in background theories. Exapmles of theories that we use and that are implem-ted in Z3).

As it has been discussed in Chapter~\ref{ch:intro}, the program may behave differently when compiled for different parallel hardware architectures. This may cause the portability bugs, the behaviour that is allowed under one architecture and forbidden under another. 
%already said:
%Although the research of weak memory models has achieved considerable success, most works remain to be rather theoretical that practical and serve mostly as tools for better understanding the concurrent nature of programs.
% The first tool that tackles the problem 
%The first work that has investigated the practical approach of modeling and verification the real-world programs with respect to weak memory models was ~\cite{Porthos17SAS}  
In this Chapter, we describe the general task of analysing the concurrent software portability
%may be stated 
as a \textit{bounded reachability} problem, which in turn can be reduced to a SAT problem~\cite{Porthos17SAS} (more precisely, to an SMT-problem).

@@

Note that the definition of portability requirements against \textit{executions} is strong enough, as it implies the portability against \textit{states} (the \textit{state-portability})~\cite{Porthos17CoRR}.
The result SMT-formula $\phi$ that encodes the portability problem should contain both encodings of control-flow $\phi_{CF}$ and data-flow $\phi_{DF}$ of the program, and assertions of both memory models: ${\phi = \phi_{CF} \land \phi_{DF} \land \phi_{\mathcal{M_T}} \land \phi_{\lnot\mathcal{M_S}}}$. If the formula is satisfiable, there exist a portability bug.
%The control-flow and data-flow encodings are standard for BMC~\cite{collavizza2006exploration}, they are described below. 
%However, encoding of memory models requires additional techniques due to recursive definitions of relations, that were proposed in~\cite{Porthos17SAS}.
